---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.retina = 3
)
```

# ggconsort

<!-- badges: start -->
<!-- badges: end -->

The goal of ggconsort is to provide convenience functions for creating CONSORT diagrams with `ggplot2`. 

## Installation

You can install the released version of ggconsort from [GitHub](https://github.com/tgerke/ggconsort) with:

``` r
devtools::install_github("tgerke/ggconsort")
```

## Example

**This paragraph needs to be modified**

In the following, we filter the `penguins` data to only those from Biscoe island. In the same `dplyr` chain, we collect the number of distinct species overall (i.e. before the filter to Biscoe) and the number of distinct species on Biscoe into a list called `counts`. In the next step, we will use `counts` to construct a basic data flow diagram.

```{r example-cohort}
library(dplyr)
library(ggconsort)
library(palmerpenguins)

penguin_cohorts <- 
  penguins %>%
  mutate(.id = row_number()) %>%
  cohort_start("Penguins observerd by Palmer Station LTER") %>%
  # Define cohorts using named expressions --------------------
  # Notice that you can use previously defined cohorts in subsequent steps
  cohort_define(
    adelie = .full %>% filter(species == "Adelie"),
    adelie_male = adelie %>% filter(sex == "male"),
    biscoe_adelie_male = adelie_male %>% filter(island == "Biscoe")
  ) %>%
  # Provide text labels for cohorts ---------------------------
  cohort_label(
    adelie = "Adelie penguins",
    adelie_male = "Adelie male penguins",
    biscoe_adelie_male = "Male Adelie penguins on Biscoe island"
  )
```

```{r example-consort}
library(ggplot2)

ggplot(data = NULL) + 
  geom_consort_arrow(
    x = 0, xend = 0, y = 50, yend = 30 
  ) + 
  geom_consort_arrow(
    x = 0, xend = 20, y = 40, yend = 40
  ) + 
  geom_consort_box(
    x = 0, y = 50, vjust = 0,
    label = glue::glue(
      '{
      cohort_count(penguin_cohorts) %>% 
        filter(cohort == ".full") %>% 
        pull(label)
      } (n = {
        cohort_count(penguin_cohorts, ".full")
      })'
    )
  ) + 
  geom_consort_box(
    x = 20, y = 40, hjust = 0,
    label = glue::glue(
      'Excluded (n = {
        cohort_count(penguin_cohorts, ".full") - 
          cohort_count(penguin_cohorts, "biscoe_adelie_male")
      })<br>
      • Not Adelie (n = {
        cohort_count(penguin_cohorts, ".full") - 
          cohort_count(penguin_cohorts, "adelie")
      })<br>
      • Not male (n = {
        cohort_count(penguin_cohorts, "adelie") - 
          cohort_count(penguin_cohorts, "adelie_male")
      })<br>
      • Not on Biscoe island (n = {
        cohort_count(penguin_cohorts, "adelie_male") - 
          cohort_count(penguin_cohorts, "biscoe_adelie_male")
      })
      '
    )
  ) +
  geom_consort_box(
    x = 0, y = 30, vjust = 1,
    label = glue::glue(
      '{
      penguin_cohorts %>% 
        summary() %>% 
        filter(cohort == "biscoe_adelie_male") %>% 
      pull(label)
      } (n = {
      penguin_cohorts %>% 
        summary() %>% 
        filter(cohort == "biscoe_adelie_male") %>% 
      pull(count)
      })'
    ), 
  ) +
  xlim(-50, 75) + 
  ylim(20, 60) + 
  theme_void()
  #theme_linedraw() #a temporary theme while in progress

```
