---
title: "PRISMA flow diagrams with ggconsort"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 6.5,
  out.width = "100%",
  fig.retina = 3
)
```

[PRISMA flow diagrams](https://www.prisma-statement.org/) document the flow of
records through a systematic review, from identification through screening to
the final set of included studies. Structurally they are close cousins of
CONSORT diagrams, so the same two-stage ggconsort workflow applies: count
cohorts while you wrangle the citation data, then lay out boxes and arrows.

This article builds a diagram in the style of the PRISMA 2020 statement for a
hypothetical systematic review.

## Citation data

In a real review, you would start from a data frame of citation records
exported from your reference manager, with one row per record and columns
that track each record's fate. Here we simulate one.

```{r citations, message = FALSE}
library(ggconsort)
library(dplyr)

citations <- tibble(id = 1:500) %>%
  mutate(
    source = if_else(id <= 412, "Databases", "Registers"),
    duplicate = id <= 102,
    passed_screen = !duplicate & id > 353,
    retrieved = passed_screen & id > 362,
    exclusion_reason = case_when(
      retrieved & id <= 409 ~ "Wrong population",
      retrieved & id <= 448 ~ "Wrong outcome",
      retrieved & id <= 469 ~ "Wrong study design",
      TRUE ~ NA_character_
    )
  )

head(citations)
```

## Stage 1: count the cohorts

Each PRISMA box is a cohort. As in a CONSORT workflow, `cohort_define()`
derives each cohort from the full set of records or from a previously defined
cohort, and `anti_join()` is a convenient way to count the records that drop
out at each step.

```{r cohorts}
review_cohorts <-
  citations %>%
  cohort_start("Records identified") %>%
  cohort_define(
    from_databases = .full %>% filter(source == "Databases"),
    from_registers = .full %>% filter(source == "Registers"),
    screened = .full %>% filter(!duplicate),
    duplicates = anti_join(.full, screened, by = "id"),
    sought = screened %>% filter(passed_screen),
    screened_out = anti_join(screened, sought, by = "id"),
    assessed = sought %>% filter(retrieved),
    not_retrieved = anti_join(sought, assessed, by = "id"),
    included = assessed %>% filter(is.na(exclusion_reason)),
    excluded_population = assessed %>% filter(exclusion_reason == "Wrong population"),
    excluded_outcome = assessed %>% filter(exclusion_reason == "Wrong outcome"),
    excluded_design = assessed %>% filter(exclusion_reason == "Wrong study design")
  ) %>%
  cohort_label(
    from_databases = "Databases",
    from_registers = "Registers",
    screened = "Records screened",
    duplicates = "Duplicate records removed",
    sought = "Reports sought for retrieval",
    screened_out = "Records excluded",
    assessed = "Reports assessed for eligibility",
    not_retrieved = "Reports not retrieved",
    included = "Studies included in review",
    excluded_population = "Wrong population",
    excluded_outcome = "Wrong outcome",
    excluded_design = "Wrong study design"
  )

review_cohorts
```

## Stage 2: lay out the diagram

The main flow runs down the left at `x = 0`, with reasons for attrition in
boxes to the right. Multi-line labels are built with `<br>` (labels are
rendered by [ggtext](https://wilkelab.org/ggtext/), so markdown and HTML
formatting work), and `hjust = 0` left-aligns the boxes that list several
items so their text hangs together — this is the same `hjust`/`vjust`
override available for any box.

```{r prisma-consort}
review_prisma <- review_cohorts %>%
  consort_box_add(
    "identified", 0, 50,
    glue::glue(
      "Records identified from:<br>
      {cohort_count_adorn(review_cohorts, from_databases)}<br>
      {cohort_count_adorn(review_cohorts, from_registers)}"
    )
  ) %>%
  consort_box_add(
    "duplicates", 30, 50,
    glue::glue(
      "Records removed before screening:<br>
      {cohort_count_adorn(review_cohorts, duplicates)}"
    ),
    hjust = 0
  ) %>%
  consort_box_add(
    "screened", 0, 40, cohort_count_adorn(review_cohorts, screened)
  ) %>%
  consort_box_add(
    "screened_out", 30, 40, cohort_count_adorn(review_cohorts, screened_out),
    hjust = 0
  ) %>%
  consort_box_add(
    "sought", 0, 30, cohort_count_adorn(review_cohorts, sought)
  ) %>%
  consort_box_add(
    "not_retrieved", 30, 30, cohort_count_adorn(review_cohorts, not_retrieved),
    hjust = 0
  ) %>%
  consort_box_add(
    "assessed", 0, 20, cohort_count_adorn(review_cohorts, assessed)
  ) %>%
  consort_box_add(
    "excluded", 30, 20,
    glue::glue(
      "Reports excluded:<br>
      • {cohort_count_adorn(review_cohorts, excluded_population)}<br>
      • {cohort_count_adorn(review_cohorts, excluded_outcome)}<br>
      • {cohort_count_adorn(review_cohorts, excluded_design)}"
    ),
    hjust = 0
  ) %>%
  consort_box_add(
    "included", 0, 10, cohort_count_adorn(review_cohorts, included)
  ) %>%
  consort_arrow_add(
    start = "identified", start_side = "bottom",
    end = "screened", end_side = "top"
  ) %>%
  consort_arrow_add(
    start = "screened", start_side = "bottom",
    end = "sought", end_side = "top"
  ) %>%
  consort_arrow_add(
    start = "sought", start_side = "bottom",
    end = "assessed", end_side = "top"
  ) %>%
  consort_arrow_add(
    start = "assessed", start_side = "bottom",
    end = "included", end_side = "top"
  ) %>%
  consort_arrow_add(
    start_x = 0, start_y = 50,
    end = "duplicates", end_side = "left"
  ) %>%
  consort_arrow_add(
    start_x = 0, start_y = 40,
    end = "screened_out", end_side = "left"
  ) %>%
  consort_arrow_add(
    start_x = 0, start_y = 30,
    end = "not_retrieved", end_side = "left"
  ) %>%
  consort_arrow_add(
    start_x = 0, start_y = 20,
    end = "excluded", end_side = "left"
  )
```

As with CONSORT diagrams, the PRISMA stage labels are ordinary ggplot layers
added on top of `geom_consort()`. Because the left-aligned boxes extend to the
right of their anchor points, we widen the x range with `xlim()` to give them
room.

```{r prisma-plot, fig.alt = "PRISMA 2020 flow diagram: 500 records identified from databases and registers, 398 screened after removing 102 duplicates, 147 reports sought, 138 assessed for eligibility, and 31 studies included in the review"}
library(ggplot2)

review_prisma %>%
  ggplot() +
  geom_consort() +
  theme_consort(margin_h = 2, margin_v = 1) +
  xlim(-32, 62) +
  ggtext::geom_richtext(
    aes(x = -28, y = y, label = label),
    data = tibble(
      # "Included" sits at 9, not 10: the included box hangs below its anchor
      y = c(50, 35, 9),
      label = c("Identification", "Screening", "Included")
    ),
    fill = "#9bc0fc", angle = 90
  )
```

At analysis time, the included studies are one `cohort_pull()` away:

```{r pull}
review_cohorts %>%
  cohort_pull(included)
```
