---
title: "PRISMA flow diagrams with ggconsort"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 6.5,
  out.width = "100%",
  fig.retina = 3
)
```

[PRISMA flow diagrams](https://www.prisma-statement.org/) document the flow of
records through a systematic review, from identification through screening to
the final set of included studies. Structurally they are close cousins of
CONSORT diagrams, so the same two-stage ggconsort workflow applies: count
cohorts while you wrangle the citation data, then lay out boxes and arrows.

This article builds a diagram in the style of the PRISMA 2020 statement for a
hypothetical systematic review.

## Citation data

In a real review, you would start from a data frame of citation records
exported from your reference manager, with one row per record and columns
that track each record's fate. Here we simulate one.

```{r citations, message = FALSE}
library(ggconsort)
library(dplyr)

citations <- tibble(id = 1:500) |>
  mutate(
    source = if_else(id <= 412, "Databases", "Registers"),
    duplicate = id <= 102,
    passed_screen = !duplicate & id > 353,
    retrieved = passed_screen & id > 362,
    exclusion_reason = case_when(
      retrieved & id <= 409 ~ "Wrong population",
      retrieved & id <= 448 ~ "Wrong outcome",
      retrieved & id <= 469 ~ "Wrong study design",
      TRUE ~ NA_character_
    )
  )

head(citations)
```

## Stage 1: count the cohorts

Each PRISMA box is a cohort. As in a CONSORT workflow, `cohort_define()`
derives each cohort from the full set of records or from a previously defined
cohort, and `anti_join()` is a convenient way to count the records that drop
out at each step.

```{r cohorts}
review_cohorts <-
  citations |>
  cohort_start("Records identified") |>
  cohort_define(
    from_databases = .full |> filter(source == "Databases"),
    from_registers = .full |> filter(source == "Registers"),
    screened = .full |> filter(!duplicate),
    duplicates = anti_join(.full, screened, by = "id"),
    sought = screened |> filter(passed_screen),
    screened_out = anti_join(screened, sought, by = "id"),
    assessed = sought |> filter(retrieved),
    not_retrieved = anti_join(sought, assessed, by = "id"),
    included = assessed |> filter(is.na(exclusion_reason)),
    excluded_population = assessed |> filter(exclusion_reason == "Wrong population"),
    excluded_outcome = assessed |> filter(exclusion_reason == "Wrong outcome"),
    excluded_design = assessed |> filter(exclusion_reason == "Wrong study design")
  ) |>
  cohort_label(
    from_databases = "Databases",
    from_registers = "Registers",
    screened = "Records screened",
    duplicates = "Duplicate records removed",
    sought = "Reports sought for retrieval",
    screened_out = "Records excluded",
    assessed = "Reports assessed for eligibility",
    not_retrieved = "Reports not retrieved",
    included = "Studies included in review",
    excluded_population = "Wrong population",
    excluded_outcome = "Wrong outcome",
    excluded_design = "Wrong study design"
  )

review_cohorts
```

## Stage 2: lay out the diagram

Each PRISMA box declares a `row` and `col` grid position: the main flow runs
down the `"main"` column, with the reasons for attrition beside it in the
`"side"` column. Arrows connect boxes by name — boxes in the same column are
joined vertically and boxes in the same row horizontally, PRISMA-style, from
box edge to box edge. The stage labels on the left are `consort_stage_add()`
badges; `row = c(2, 4)` centers "Screening" across those rows. Multi-line
labels are built with `<br>` (labels are rendered by
[gridtext](https://wilkelab.org/gridtext/), so markdown and HTML formatting
work).

ggconsort measures every box when the plot is drawn and computes the spacing
to fit the figure, so no coordinates are needed anywhere.

```{r prisma-consort}
review_prisma <- review_cohorts |>
  consort_box_add(
    "identified", row = 1, label = glue::glue(
      "Records identified from:<br>
      {cohort_count_adorn(review_cohorts, from_databases)}<br>
      {cohort_count_adorn(review_cohorts, from_registers)}"
    )
  ) |>
  consort_box_add(
    "duplicates", row = 1, col = "side", label = glue::glue(
      "Records removed before screening:<br>
      {cohort_count_adorn(review_cohorts, duplicates)}"
    )
  ) |>
  consort_box_add(
    "screened", row = 2, label = cohort_count_adorn(review_cohorts, screened)
  ) |>
  consort_box_add(
    "screened_out", row = 2, col = "side",
    label = cohort_count_adorn(review_cohorts, screened_out)
  ) |>
  consort_box_add(
    "sought", row = 3, label = cohort_count_adorn(review_cohorts, sought)
  ) |>
  consort_box_add(
    "not_retrieved", row = 3, col = "side",
    label = cohort_count_adorn(review_cohorts, not_retrieved)
  ) |>
  consort_box_add(
    "assessed", row = 4, label = cohort_count_adorn(review_cohorts, assessed)
  ) |>
  consort_box_add(
    "excluded", row = 4, col = "side", label = glue::glue(
      "Reports excluded:<br>
      • {cohort_count_adorn(review_cohorts, excluded_population)}<br>
      • {cohort_count_adorn(review_cohorts, excluded_outcome)}<br>
      • {cohort_count_adorn(review_cohorts, excluded_design)}"
    )
  ) |>
  consort_box_add(
    "included", row = 5, label = cohort_count_adorn(review_cohorts, included)
  ) |>
  consort_arrow_add(start = "identified", end = "screened") |>
  consort_arrow_add(start = "screened", end = "sought") |>
  consort_arrow_add(start = "sought", end = "assessed") |>
  consort_arrow_add(start = "assessed", end = "included") |>
  consort_arrow_add(start = "identified", end = "duplicates") |>
  consort_arrow_add(start = "screened", end = "screened_out") |>
  consort_arrow_add(start = "sought", end = "not_retrieved") |>
  consort_arrow_add(start = "assessed", end = "excluded") |>
  consort_stage_add("Identification", row = 1, angle = 90) |>
  consort_stage_add("Screening", row = c(2, 4), angle = 90) |>
  consort_stage_add("Included", row = 5, angle = 90)
```

```{r prisma-plot, fig.alt = "PRISMA 2020 flow diagram: 500 records identified from databases and registers, 398 screened after removing 102 duplicates, 147 reports sought, 138 assessed for eligibility, and 31 studies included in the review"}
library(ggplot2)

review_prisma |>
  ggplot() +
  geom_consort() +
  theme_consort()
```

At analysis time, the included studies are one `cohort_pull()` away:

```{r pull}
review_cohorts |>
  cohort_pull(included)
```
